---
title: "ASV rare species analysis"
author: "xyz"
date: "2021/4/28"
output: html_document
---

### filter rare species

rare (<1% relative abundance) species. Albertsen, M., Hugenholtz, P., Skarshewski, A. et al. Genome sequences of rare, uncultured bacteria obtained by differential coverage binning of multiple metagenomes. Nat Biotechnol 31, 533–538 (2013). https://doi.org/10.1038/nbt.2579

```{r Rare ASV cutoff and ratio}
ASV <- read.csv("../table/ASVwithTaxByNaive_bayesSortByOneMore.csv")
totalAbundance <- ASV$totalPercent
maxAbundance <- max(totalAbundance)
ASVCount <- nrow(ASV)
# Rare species cut off is from 0 to 0.1%
RareSpeciesCutoff <- seq(0, 0.1, length.out = 1000)
RareSpeciesRatio <-
  sapply(RareSpeciesCutoff, function(x) {
    sum(totalAbundance < x) / ASVCount * 100
  }, simplify = T)
RareSpeciesDF <-
  data.frame(RareSpeciesCutoff = RareSpeciesCutoff, RareSpeciesRatio = RareSpeciesRatio)

p <-
  ggplot(data = RareSpeciesDF,
         mapping = aes(x = RareSpeciesCutoff, y = RareSpeciesRatio)) + geom_line() +
  xlab("Rare ASV cutoff (%)") +
  ylab("Rare ASV ratio (%)") +
  theme(text = element_text(size = 20))

xintercept <- 0.01
yintercept <- sum(sum(totalAbundance < 0.01) / ASVCount) * 100
# more than 90% ASVs' abundances are less than 0.01%
# 91.33
round(yintercept, 2)
noteLineDf <-
  data.frame(
    x1 = c(xintercept, -Inf),
    x2 = c(xintercept,xintercept),
    y1 = c(-Inf, yintercept),
    y2 = c(yintercept, yintercept)
  )
p <-
  p + geom_segment(aes(
    x = x1,
    y = y1,
    xend = x2,
    yend = y2,
  ), data = noteLineDf,linetype = 4)+
  scale_x_continuous(breaks = c(0,xintercept ,0.025, 0.05, 0.075,0.1))+
  scale_y_continuous(breaks = c(0,25 ,50,75,round(yintercept,1),100))
p + ggsave("../figure/Rare ASV cutoff and ratio.pdf",
           width = 8,
           height = 6)
```

The cumulative distribution and frequency stats of abundance of rare ASV

In this paper, they declaim abundance fluctuations are gamma distributed.

Grilli, J. Macroecological laws describe variation and diversity in microbial communities. Nat Commun 11, 4743 (2020). https://doi.org/10.1038/s41467-020-18529-y

```{r The cumulative distribution and frequency stats of abundance of rare ASV}
library(scales)
Abundance <- data.frame(Abundance = totalAbundance[totalAbundance<0.01])
p <-
  ggplot(data = Abundance,
         aes(x = Abundance))
# xlim(0,0.1)+
p2 <- p + stat_ecdf(geom = "step", pad = FALSE) +
  scale_y_continuous(labels = percent, name = "ASV ratio (%)") +
  xlab("ASV abundance (%)")
# it is same as Rare ASV cutoff and ratio.pdf
p2

# frequency stats

p3 <- p + geom_histogram(binwidth = 0.01 / 50)
p3 <- p3 + xlab("ASV abundance (%)") +
  ylab("ASV counts") +
  theme(text = element_text(size = 20))

p3 + ggsave("../figure/Rare ASV frequency stats.pdf",
           width = 10,
           height = 6)
```

```{r abundance acculation of ASV}
rankedTotalAbundance<-sort(totalAbundance)
accumulateAbundance<-accumulate(rankedTotalAbundance,`+`)
accumulateAbundanceTable<-data.frame(accumulateAbundance=accumulateAbundance,
                                     ASVratio=((1:ASVCount)/ASVCount*100))

# 94.38% ASV's abundance<0.01, of which the sum of abundance is 20.82%
notePoint<-accumulateAbundanceTable[sum(rankedTotalAbundance<0.01),]
xintercept<-notePoint$ASVratio
yintercept<-notePoint$accumulateAbundance
noteLineDf <-
  data.frame(
    x1 = c(xintercept, -Inf),
    x2 = c(xintercept,xintercept),
    y1 = c(-Inf, yintercept),
    y2 = c(yintercept, yintercept)
  )
p <-
  ggplot(data = accumulateAbundanceTable,
         mapping = aes(x = ASVratio, y = accumulateAbundance)) + geom_line() +
  xlab("ASV ratio (%)")+
  ylab("Accumulate Abundance (%)") +
  theme(text = element_text(size = 20))
p <-
  p + geom_segment(aes(
    x = x1,
    y = y1,
    xend = x2,
    yend = y2,
  ), data = noteLineDf,linetype = 4)+
  scale_x_continuous(breaks = c(0,20 ,50, 75, round(xintercept,1),100))+
  scale_y_continuous(breaks = c(0,round(yintercept,1),25 ,50,75,100))
p + ggsave("../figure/Rare ASV abundance acculation.pdf",
           width =12,
           height = 6)
```

### resample cutoff step 0.001

```{r normalize}
library(DESeq2)
ASVForNormalization <- ASV[, 2:220]
coldata <- data.frame(row.names = colnames(ASVForNormalization))
# ~ 1 can be used for no design
dds <- DESeqDataSetFromMatrix(countData = ASVForNormalization,
                              colData = coldata,
                              design =  ~ 1)
dds <- estimateSizeFactors(dds, type = 'poscounts')
# sizeFactors(dds)
normalizedASV <- counts(dds, normalized = TRUE)
normalizedASV <- data.frame(normalizedASV)
```

#### resample

```{r}
library(dplyr)

getRareSpeciesRatio <- function(ASV, cutoffs) {
  totalAbundance <- rowSums(ASV)
  # remove ASV with zero abundance
  totalAbundance <- totalAbundance[totalAbundance > 0]
  totalCount <- sum(totalAbundance)
  totalAbundance <- totalAbundance / totalCount * 100
  ASVCount <- length(totalAbundance)
  SpeciesRatio <-
    sapply(cutoffs, function(cutoff) {
      sum(totalAbundance < cutoff) / ASVCount * 100
    }, simplify = T)
  return(SpeciesRatio)
}

# cutoff is from 0.01 to 1
cutoffs <- seq(0.01, 1, length.out = 100)
# resample 1~100 sample, each time 10
ns <- rep(1:100, each = 10)
seeds <- 1:length(ns)

reSample <- function(seed, n, pool) {
  set.seed(seed)
  sample(pool, n)
}
reSampledColIndex <-
  mapply(reSample, seeds, ns, MoreArgs = list(pool = 1:219))

reSampledRareSpeciesRatio <-
  mapply(
    function(colIndex, cutoffs) {
      resampledASV <- normalizedASV[colIndex]
      ratio <- getRareSpeciesRatio(resampledASV, cutoffs)
      return(ratio)
    },
    reSampledColIndex,
    MoreArgs = list(cutoffs = cutoffs),
    SIMPLIFY = T
  )

reSampledResults <- data.frame(
  ns = rep(ns, each = length(cutoffs)),
  seeds = rep(seeds, each = length(cutoffs)),
  cutoffs = rep(cutoffs, time = length(ns)),
  reSampledRareSpeciesRatio = c(reSampledRareSpeciesRatio)
)
reSampledResultsSummary <- group_by(reSampledResults, ns, cutoffs)
reSampledResultsSummary <-
  summarise(
    reSampledResultsSummary,
    SpeciesRatioMean = mean(reSampledRareSpeciesRatio),
    SpeciesRatioSD = sd(reSampledRareSpeciesRatio)
  )
```
#### render 3D plot in plot3D

[plot3D](http://www.sthda.com/english/wiki/impressive-package-for-3d-and-4d-graph-r-software-and-data-visualization#change-the-position-of-the-legend)

```{r}
library("plot3D")
`Replicate count` = reSampledResultsSummary$ns
`Rare species cutoff` = reSampledResultsSummary$cutoffs
`Rare species ratio` = reSampledResultsSummary$SpeciesRatioMean
pdf(
  "../figure/replicate count 0-100,cutoff 0.01-0.1, step 0.01 and Rare species ratio 3D.pdf",
  20,
  9
)
scatter3D(
  `Replicate count`,
  `Rare species cutoff`,
  `Rare species ratio`,
  colvar = reSampledResultsSummary$SpeciesRatioSD,
  # horizontal angel
  theta = -45,
  # vertical angel
  phi = 0,
  bty = "g",
  pch = 20,
  # point size
  cex = 1,
  # font size
  cex.axis = 1.5,
  cex.lab = 1.5,
  ticktype = "detailed",
  xlab = "Replicate count",
  ylab = "Rare species cutoff (%)",
  zlab = "Rare species ratio (%)",
  clab = "Species ratio SD (%)",
  # legend parameter, font.clab to set font plain or bold is invalid
  colkey = list(
    shift = 0,
    side = 4,
    length = 0.5,
    cex.clab = 1.5,
    font.clab = 1
  )
)
dev.off()
```

#### tieplot in ggplot2

```{r}
library(ggplot2)
df <- reSampledResultsSummary
p <- ggplot(df) +
  geom_tile(aes(
    ns,
    cutoffs,
    fill = SpeciesRatioMean,
    width = 1,
    height = 0.01
  )) +
  scale_fill_gradientn(colours = topo.colors(15)) +
  theme_bw() +
  ylab("Rare species cutoff (%)") +
  xlab("Replicate count") +
  # change legend title
  labs(fill = "Rare species ratio (%)") +
  theme(text = element_text(size = 20))
p + ggsave(
  "../figure/replicate count 0-100,cutoff 0.01-0.1, step 0.01 and Rare species ratio.pdf",
  width = 12,
  height = 9
)

p <- ggplot(df) +
  geom_tile(aes(
    ns,
    cutoffs,
    fill = SpeciesRatioSD,
    width = 1,
    height = 0.01
  )) +
  scale_fill_gradientn(colours = topo.colors(15)) +
  theme_bw() +
  ylab("Rare species cutoff (%)") +
  xlab("Replicate count") +
  # change legend title
  labs(fill = "Rare species ratio SD (%)") +
  theme(text = element_text(size = 20))
p + ggsave(
  "../figure/replicate count 0-100,cutoff 0.01-0.1, step 0.01 and Rare species ratio SD.pdf",
  width = 12,
  height = 9
)
```

